---
title: "Foundation Models for Multi-Study Conjoint Learning"
author: "strategize package authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundation Models for Multi-Study Conjoint Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

CONDA_AVAILABLE <- FALSE

if (identical(Sys.getenv("STRATEGIZE_RUN_VIGNETTE"), "true")) {
  CONDA_AVAILABLE <- tryCatch({
    if (requireNamespace("reticulate", quietly = TRUE)) {
      conda_list <- reticulate::conda_list()
      "strategize" %in% conda_list$name || "strategize_env" %in% conda_list$name
    } else {
      FALSE
    }
  }, error = function(e) FALSE)
}

if (!CONDA_AVAILABLE && interactive()) {
  message(
    "Note: Foundation-model vignette code is not executed by default. ",
    "To run code: Sys.setenv(STRATEGIZE_RUN_VIGNETTE='true') before rendering."
  )
}
```

# Overview

The foundation-model workflow in `strategize` is meant for settings where you
have many conjoint experiments and want to learn one reusable neural
representation across them, then adapt that representation to a target study.

The workflow has two stages:

1. **Pooled foundation training** across multiple experiments with
   `fit_conjoint_foundation_model()`.
2. **Bayesian study adaptation** into the existing neural outcome model with
   `adapt_conjoint_foundation_model()`.

This differs from the standard `strategize()` workflow, which focuses on
single-study optimization. Here the goal is to learn transferable
representations across many studies first, then specialize.

# Backend Setup

Foundation-model training uses the same JAX backend as the package's neural
outcome model. Run this once before fitting:

```{r backend, eval=FALSE}
library(strategize)

build_backend(conda_env = "strategize_env")
```

After that, you can point training and adaptation calls at the same conda
environment with `conda_env = "strategize_env"`.

# The `experiments` Contract

`fit_conjoint_foundation_model()` expects a list of per-study specifications.
Each study must provide:

- `experiment_id`: unique study identifier
- `Y`: outcome vector
- `W`: factor matrix or data frame

The core shape rule is:

```{r shape-rule, eval=FALSE}
length(Y) == nrow(W)
```

Optional fields include:

- `mode`: `"auto"`, `"pairwise"`, or `"single"`
- `pair_id`, `profile_order`: required for pairwise studies
- `X`: optional numeric covariates aligned row-wise to `W`
- `respondent_id`, `respondent_task_id`
- `likelihood`: `"auto"`, `"bernoulli"`, `"categorical"`, or `"normal"`
- `n_outcomes`: categorical class count when needed
- `names_list`, `p_list`
- `canonical_factor_id`, `canonical_level_id`

## Pairwise Requirements

For forced-choice conjoint studies:

- `pair_id` must be supplied
- each `pair_id` must appear exactly twice
- `profile_order` is optional but recommended and usually takes values `1` and `2`

## Outcome Families

The foundation workflow can ingest:

- Bernoulli outcomes
- categorical outcomes
- Gaussian outcomes

`strategize` does **not** force all of them into one universal output head.
Instead, the returned foundation object stores separate internal groups for
compatible families defined by:

```{r grouping, eval=FALSE}
(mode, likelihood, n_outcomes)
```

That means a mix of pairwise Bernoulli and single-study Gaussian experiments
will produce one `conjoint_foundation_model`, but with separate internal pooled
fits.

# Minimal Pooled Training Example

The simplest useful example is two pairwise binary studies that share some
factors and differ on others.

```{r pooled-example-data, eval=FALSE}
library(strategize)

study_a <- list(
  experiment_id = "study_a",
  Y = c(1, 0, 0, 1),
  W = data.frame(
    price = c("Low", "Low", "High", "High"),
    message = c("Jobs", "Taxes", "Jobs", "Taxes")
  ),
  pair_id = c(1, 1, 2, 2),
  profile_order = c(1, 2, 1, 2),
  canonical_factor_id = c(
    price = "price",
    message = "message"
  )
)

study_b <- list(
  experiment_id = "study_b",
  Y = c(0, 1, 1, 0),
  W = data.frame(
    price = c("Low", "High", "Low", "High"),
    message = c("Jobs", "Jobs", "Taxes", "Taxes"),
    messenger = c("Local", "Local", "National", "National")
  ),
  pair_id = c(1, 1, 2, 2),
  profile_order = c(1, 2, 1, 2),
  canonical_factor_id = c(
    price = "price",
    message = "message",
    messenger = "messenger"
  )
)
```

Fit the pooled foundation model:

```{r pooled-fit, eval=CONDA_AVAILABLE}
foundation_fit <- fit_conjoint_foundation_model(
  experiments = list(study_a, study_b),
  foundation_control = list(
    neural_mcmc_control = list(
      ModelDims = 32L,
      ModelDepth = 1L,
      subsample_method = "batch_vi",
      uncertainty_scope = "output",
      svi_steps = 100L
    )
  ),
  conda_env = "strategize_env"
)
```

Inspect the internal pooled groups:

```{r pooled-groups, eval=CONDA_AVAILABLE}
names(foundation_fit$groups)
```

In this example you should see one pairwise Bernoulli group because both
studies are compatible.

# How Schema Sharing Works

Schema sharing is conservative by default.

## Canonical ids force sharing

If two studies both use:

```{r canonical-sharing, eval=FALSE}
canonical_factor_id = c(price = "price")
```

then the pooled model treats those factor slots as one shared factor identity.

The same logic applies to `canonical_level_id`, which can force specific levels
to share identities across studies.

## Raw label equality does not force sharing

If two studies both contain a level labeled `"Moderate"` but you do not provide
canonical ids, `strategize` keeps them local to their respective studies. This
avoids accidental merging of semantically different labels.

## Missing factors are allowed

When one study has a factor that another study does not, the pooled schema still
contains that slot. Rows from studies that do not contain the factor route that
slot to the holdout row during pooled training.

# Optional Text Semantics

Text embeddings are available as optional side information, not as the primary
identity mechanism.

If you want to add a small semantic side channel, provide a
`foundation_control$text_embedding_fn` that maps character inputs to fixed-width
numeric embeddings.

For example:

```{r text-embeddings, eval=FALSE}
text_embedding_fn <- function(x) {
  x <- as.character(x)
  cbind(
    nchar = nchar(x),
    has_space = 1 * grepl(" ", x),
    has_upper = 1 * grepl("[A-Z]", x)
  )
}

foundation_fit <- fit_conjoint_foundation_model(
  experiments = list(study_a, study_b),
  foundation_control = list(
    text_embedding_fn = text_embedding_fn
  )
)
```

These embeddings are appended as pooled semantic side features. They do not
override canonical ids and they do not force schema merges on their own.

# Mixed-Family Behavior

If you mix incompatible experiment families, one foundation object is still
returned, but it stores multiple internal groups.

For example:

```{r mixed-families, eval=FALSE}
single_normal <- list(
  experiment_id = "single_normal",
  Y = c(0.2, 0.5, 0.7, 0.4),
  W = data.frame(
    policy = c("A", "B", "C", "A"),
    messenger = c("Local", "National", "Local", "National")
  ),
  mode = "single",
  likelihood = "normal"
)

foundation_fit <- fit_conjoint_foundation_model(
  experiments = list(study_a, study_b, single_normal)
)

names(foundation_fit$groups)
```

You should expect separate entries such as a pairwise Bernoulli group and a
single-study Gaussian group.

# Adaptation to a Target Study

Once a foundation model is available, adapt it to a target study with
`adapt_conjoint_foundation_model()`.

```{r adaptation, eval=CONDA_AVAILABLE}
target_study <- data.frame(
  price = c("Low", "High", "Low", "High"),
  message = c("Jobs", "Jobs", "Taxes", "Taxes")
)

adapted_fit <- adapt_conjoint_foundation_model(
  foundation_model = foundation_fit,
  Y = c(1, 0, 0, 1),
  W = target_study,
  mode = "pairwise",
  pair_id = c(1, 1, 2, 2),
  profile_order = c(1, 2, 1, 2),
  experiment_id = "target_study",
  canonical_factor_id = c(price = "price", message = "message"),
  conda_env = "strategize_env"
)
```

The return value is a regular `strategic_predictor`, so prediction works through
the standard interface:

```{r adaptation-predict, eval=CONDA_AVAILABLE}
predict(
  adapted_fit,
  newdata = list(
    W = target_study,
    pair_id = c(1, 1, 2, 2),
    profile_order = c(1, 2, 1, 2)
  )
)
```

## What adaptation does

Adaptation does **not** replace the current neural model family. Instead it:

1. matches the target study to a compatible foundation group
2. warm-starts the local neural parameters from the pooled fit when possible
3. runs the package's current Bayesian neural fit on the target study

Matched factors and levels inherit warm-started embeddings. Unmatched local
schema elements fall back to local initialization unless you request strict
schema matching.

# Save and Load

Foundation objects can be cached to disk:

```{r save-load, eval=CONDA_AVAILABLE}
save_conjoint_foundation_bundle(
  file = "foundation_bundle.rds",
  foundation_model = foundation_fit,
  overwrite = TRUE
)

foundation_fit_reloaded <- load_conjoint_foundation_bundle(
  file = "foundation_bundle.rds",
  preload_params = FALSE
)
```

`preload_params = FALSE` is usually the right default when you only need
metadata or plan to adapt later.

# Common Failure Modes

The most common setup errors are:

- `pair_id` missing in pairwise mode
- `pair_id` values that do not appear exactly twice
- `length(Y) != nrow(W)`
- non-numeric columns in `X`
- forcing `likelihood = "categorical"` with an `n_outcomes` that does not match
  the observed classes in the study
- attempting adaptation when no compatible foundation group exists
- assuming raw label equality implies cross-study sharing when no canonical ids
  were supplied

# Runtime Notes

Foundation-model training is substantially heavier than the package's GLM path.
For exploratory work, start with a shallow configuration:

```{r runtime-notes, eval=FALSE}
foundation_control = list(
  neural_mcmc_control = list(
    ModelDims = 32L,
    ModelDepth = 1L,
    subsample_method = "batch_vi",
    uncertainty_scope = "output",
    svi_steps = 100L
  )
)
```

Then scale up only after the data contract and schema sharing rules are stable.
